# Structure of data
We will use `mtcars` to illustrate the structure of data.

<br>

## Classify variables

- Recall there are 2 main types of variables: qualitative and quantitative, which R calls `Factor` and `num`.

- View the data.
```{r,results='hide'}
mtcars
head(mtcars)
tail(mtcars, n=3)
help(mtcars)
```

- Calculate the `dim`ensions of the data set.
```{r,results='hide'}
dim(mtcars)
```
This means that there are 32 rows (the types of cars) and 11 variables (properties of the cars).

<br>

- List the `names` of the variables.
```{r,results='hide'}
names(mtcars)
```

<br>

- See how R has classified the variables by viewing the `str`ucture of the data.
```{r,eval=F}
str(mtcars)
```

where 'num' is a quantitative (numerical) variable and 'Factor' is a qualitative (categorical) variable.

<br>

## Isolate a variable

- Choose one variable from the data frame by using `DataName$VariableName` and store the result in a vector.

```{r,eval=F}
mpg= mtcars$mpg
```

Note that RStudio has code completion, so will auto-predict your commands. When you type `mtcars$`, the names of the all the variables will come up.

<br>

- See the `class`ification of 1 variable.
```{r,eval=F}
class(mpg)
str(mpg)
```

<br>

- See the `length` of 1 variable.
```{r,eval=F}
length(mpg)
```

<br>

- Calculate the `sum` of a (quantitative) variable.
```{r,eval=F}
sum(mpg)
```

- If at any command you get the answer NA, it means that you need to specify what to do with missing values. 
See <i class="fa fa-link"></i>[Resource](https://stat.ethz.ch/R-manual/R-devel/library/base/html/sum.html) on how to solve this.

<br>

- `Sort` the data in increasing order.
```{r,eval=F}
sort(mpg)
```

<br>

- [Work out](https://stat.ethz.ch/R-manual/R-devel/library/base/html/sort.html) how to sort the data in decreasing order.

```{r,eval=F}
sort(mpg, decreasing = T)
```

<br>

- Sum the 5 lowest values of the variable.
```{r,eval=F}
sum(sort(mpg)[1:5])
```

<br>

## Select subset 

- Pick the 1st and 5th elements of the vector `mpg`

```{r,eval=F}
mpg[1]
mpg[5]
mpg[c(1,5)]
mtcars$mpg[c(1,5)]
mtcars[1,1]  
mtcars[5,1]   #mpg is 1st column.
```

<br>

## Change classification

- You may not agree with R's initial classification, and want to change it.

```{r,results='hide'}
str(mtcars)
```

- For example, note that the number of carburetors `carb` is classified as `num`. Reclassify `carb` as a `factor`.
```{r,results='hide'}
class(mtcars$carb)
carbF = factor(mtcars$carb)
class(carbF)
```

- To change from a `factor` to a `num`:
```{r,results='hide'}
ageCanVote = factor(setNames(c(16, 18, 18, "Unknown"), c("Austria", "Australia", "Afghanistam", "Zambia")))
as.numeric(ageCanVote)  # This is a mistake, as it converts to the rank of the factor level
as.numeric(as.character(ageCanVote))  # This converts properly
```

Note: 
(1) The warning message is not a problem - it is just alerting you to the introduction of NAs.
(2) The mistake above if you just use `as'numeric()`.

<br><br>