# Graphical Summaries {#graphicalsummaries}

The graphical summary must match up with the type of variable(s).

|Variable|Type of summary|
|----------|------------------------|
|1 Qualitative|(Single) Barplot|
|1 Quantitative|Histogram or (Single) Boxplot|
|2 Qualitative|Double Barplot, Mosaicplot|
|2 Quantitative|Scatterplot|
|1 Quantitative, 1 Qualitative|Double Boxplot|

<br>

## Barplot 

A barplot is used for qualitative variables.

- Which variables are qualitative in `mtcars`?

-  Produce a **single** `barplot` of the gears.
```{r,fig.show='hide'}
barplot(mtcars$gear)
```
Notice this is not useful! This is because R has classified `gear` as a quantitative variable.

- Instead, first summarise `gears` into a table as follows.
```{r,fig.show='hide'}
# Produce frequency table
table(mtcars$gear)

# produce barplot (from frequency table)
counts = table(mtcars$gear)
barplot(counts)
```

- Now customise the barplot.
```{r,fig.show='hide'}
help(barplot)
barplot(counts, names.arg=c("3 Gears","4 Gears","5 Gears"),col="lightblue")
```

- Make the names of bars perpendicular to axis.
```{r,fig.show='hide'}
par(las=2) 
barplot(counts, names.arg=c("3 Gears","4 Gears","5 Gears"),col="lightblue")
```

- Now consider 2 qualitative variables: `gear` and `cyl`. Produce a **double** barplot by **faceting** or **filtering** the barplot of `gear` by `cyl`.
```{r,fig.show='hide'}
counts1 = table(mtcars$cyl, mtcars$gear)
barplot(counts1,names.arg=c("3 Gears","4 Gears","5 Gears"),col=c("lightblue","lightgreen","lightyellow"),legend = rownames(counts1))
barplot(counts1,names.arg=c("3 Gears","4 Gears","5 Gears"),col=c("lightblue","lightgreen","lightyellow"),legend = c("4 cyl","6 cyl","8 cyl"))
barplot(counts1, names.arg = c("3 Gears", "4 Gears", "5 Gears"), col = c("lightblue", 
    "lightgreen", "lightyellow"), legend = c("4 cyl", "6 cyl", "8 cyl"), beside = TRUE)
```
What do you learn?

<br>

## Histogram

A histogram is used for quantitative variables.

- Which variables are quantitative in `mtcars`?

- Produce a `hist`ogram of the weights.
```{r,fig.show='hide'}
hist(mtcars$wt)
```

- Produce a probability `hist`ogram of the weights. What is the difference? Why do the 2 histograms have an identical shape here? 
```{r,fig.show='hide'}
hist(mtcars$wt,freq=F)
```

- In this course, we will consider the probability histogram (2nd one) which means that the total area of the histogram is 1.

- What does the histogram tell us about weights of the cars?

- To see what customisations for `hist` are available, use `help`.
```{r,results='hide'}
help(hist)
```

- Try this `hist`ogram of the weights.
```{r,fig.show='hide'}
hist(mtcars$wt, br=seq(0,6,by=0.5), freq=F, col="lightgreen",xlab="weight of cars (1000 lbs)",main="Histogram of Weights of Cars US 1973-74")
```

- Experiment with the customisations to see how they work.

- Try a `hist`ogram of the gross horsepower. What do you learn?

- Produce a `hist`ogram of mpg. What do you learn?

<br>

## Boxplot

A boxplot is another summary for quantitative variables.

- Produce a **single** boxplot for the weights of cars.
```{r,fig.show='hide'}
boxplot(mtcars$wt)
```

- Produce a horizontal boxplot.
```{r,fig.show='hide'}
boxplot(mtcars$wt, horizontal = T)
```

Which orientation do you prefer?

Compare to the histogram of weights above:  what different features are highlighted by a boxplot?

- Customise your boxplot.
```{r,results='hide'}
help(boxplot)
```

- Now consider dividing the weights (qualitative) by cylinders (qualitative). Produce a **double boxplot**, by **filtering** or **faceting** `wt` by `cyl`.

```{r,fig.show='hide'}
boxplot(mtcars$wt~mtcars$cyl)
boxplot(mtcars$wt~mtcars$cyl, names=c("4 cyl", "6 cyl","8 cyl"),ylab="Weight of cars (1000 lbs)")
```
What do you learn about the weights of cars?
<i class="fa fa-link"></i>[Car Weight - see page 6](http://web.mit.edu/sloan-auto-lab/research/beforeh2/files/MacKenzie%20Zoepf%20Heywood%20Car%20Weight%20Trends%20-%20IJVD.pdf)

- Try facetting the weights by another qualitative variable. 

<br>

## Mosaicplot

A mosaic plot visualises 2 qualitative variables.
```{r, eval = T}
counts2 = table(mtcars$gear, mtcars$am)  # Produces contingency table
plot(counts2) # Produces mosaic plot from contingency table
```

<br>

## Scatterplot

A scatter plot considers the relationship between 2 quantitative variables.

- What does this plot tell us?

```{r,fig.show='hide'}
plot(mtcars$wt,mtcars$mpg)
```

- Customise your scatterplot.
You can change the [plotting symbols](http://www.statmethods.net/advgraphs/parameters.html).
```{r,fig.show='hide'}
plot(mtcars$wt,mtcars$mpg, xlab="Car Weight", ylab="Miles per Gallon",col="darkred",pch=19)
```

- Add a line of best fit. 
```{r,eval=F}
abline(lm(mtcars$mpg~mtcars$wt))
```
We'll expore this further in Section 6.

- We can compare pairs of multiple quantitative variables using `plot` or `pairs`.
```{r,fig.show='hide'}
plot(mtcars)
pairs(~mpg+disp+drat+wt,data=mtcars)
pairs(mtcars)
```
Which variables seem to be related?

<br>

<br><br>

## ggplot

All the previous (base R) plots can be done in `ggplot`, allowing much greater customisation.

- First, download the [package](http://tidyverse.org/)`ggplot` into RStudio. This is a one off command.
```{r,eval=F}
install.packages("ggplot2")
```

- Each time you open RStudio, load the `ggplot2` package
```{r, message=F, warning=F}
library(ggplot2)
```

<br>

### Barplot

- Produce a **single** barplot.
```{r,fig.show='hide'}
# mtcars data
p = ggplot(mtcars, aes(x=factor(cyl)))  # Select the mtcars data, and focus on cyl as factor (qualitative) on x axis
p + geom_bar()  # Produce a barplot

# mpg data
p1 = ggplot(mpg, aes(class)) # Select the mpg data, and focus on class as x axis
p1 + geom_bar() # (1) Produce a barplot
p1 + geom_bar(aes(weight = displ)) # (2) Produce a barplot with counts from displacement variable
```

- Produce a **double** barplot.
```{r,fig.show='hide'}
p1 + geom_bar(aes(fill = drv)) # (3) Produce a (double) barplot divided by the drive variable
p1 +
 geom_bar(aes(fill = drv), position = position_stack(reverse = TRUE)) +
 coord_flip() +
 theme(legend.position = "top")  # (4) Customising (3)
```

<br><br>

### Histogram
Produce a `hist`ogram of the weights.
```{r,eval=F}  
p = ggplot(data=mtcars, aes(x=wt))  # Select the mtcars data, and focus on wt (quantitative) on x axis
p + geom_histogram(aes(y=..density..),binwidth=0.5)  
+ xlab('Weight')+ylab('Density') # Produce a histogram with x and y axis labels
```

Using `aes(y=..density..)` turns a raw histogram into a probability histogram.

<br>

### Boxplot

- Produce a **single** boxplot.
```{r,fig.show='hide'}
p = ggplot(data=mtcars, aes(x="", y=wt))  # Select the mtcars data, and focus on wt (quantitative) on y axis (with no filtering on x axis)
p + geom_boxplot() # Produce a boxplot
```

- Produce a **double** boxplot.
```{r,fig.show='hide'}
p = ggplot(data=mtcars, aes(x=factor(cyl),y=wt))  # Select the mtcars data, and focus on wt (quantitative) on y axis and cyl (qualitative) on x axis
p + geom_boxplot() # Produce a boxplot, of wt filtered by cyl
```

<br>

- `geom_jitter` plots the points with a small amount of random noise. We use it to investigate over-plotting in small data sets.
```{r,fig.show='hide'}
p = ggplot(data=mtcars, aes(x=factor(cyl),y=wt))
p + geom_boxplot() + geom_jitter()
```

<br>
What do the following customisations do? 
```{r,fig.show='hide', warning=F, message=F}
p = ggplot(data=mtcars, aes(x=factor(cyl),y=wt))
p + geom_boxplot() + coord_flip()
p + geom_boxplot(notch = TRUE)
p + geom_boxplot(outlier.colour = "green", outlier.size = 3)
```

<br>
Now consider dividing the weights by another qualitative variable.

```{r,fig.show='hide'}
p + geom_boxplot(aes(fill = factor(cyl)))
p + geom_boxplot(aes(fill = factor(am)))
```

<br>

### Scatterplot

- Produce a scatterplot.

```{r,fig.show='hide'}
p = ggplot(mtcars, aes(wt, mpg)) # Select the mtcars data, and focus on wt (quantitative) on x axis and mpg (quantiative) on y axis
p + geom_point()  # Produce a scatterplot of mpg vs wt
```

```{r,fig.show='hide'}
p + geom_point(aes(colour = factor(cyl)))  # Colour the points by cyl (qualitative)
```

<br>

- Notice what these customisations do. 

```{r,fig.show='hide'}
p + geom_point(aes(shape = factor(cyl)))  # Shape the points by cyl (qualitative)
p + geom_point(aes(shape = factor(cyl))) + scale_shape(solid = FALSE) 
p + geom_point(aes(size = qsec))  # Size the points by qsec (qualitative)
p + geom_point(aes(colour = cyl)) + scale_colour_gradient(low = "blue")  # Colour the points by cyl (quantitative)
```

<br>

## plotly **

`plotly` or `plot.ly` is included in the `ggplot2 package and allows you to have offline interactive tools.

- Produce this scatter plot. Notice what happens when you hover over the data points.

```{r,eval = T,message=F, warning=F}
library("plotly")
p1 = plot_ly(mtcars, x = ~mpg, y = ~wt, type="scatter") 
p1
```

- Now add another variable.

```{r,eval = T,message=F, warning=F}
p2 = plot_ly(mtcars, x = ~mpg, y = ~wt, symbol = ~cyl, type="scatter") 
p2
p3 = plot_ly(mtcars, x = ~mpg, y = ~wt, color = ~cyl, type="scatter") 
p3
p4 = plot_ly(mtcars, x = ~mpg, y = ~wt, color = ~factor(cyl), type="scatter") 
p4
```

- Try a boxplot.

```{r,eval = T,message=F, warning=F}
library("plotly")
p5 = plot_ly(mtcars, x = ~mpg, y = ~factor(cyl), type='box') 
```

- Here's the [cheatsheet](xhttps://images.plot.ly/plotly-documentation/images/r_cheat_sheet.pdf?_ga=2.230192159.2094081876.1551756856-1656530949.1551756856).
- More examples [here](https://www.r-graph-gallery.com/121-manage-colors-with-plotly/).
- Guide to [plotly](https://plotly-book.cpsievert.me/).

<br>

