# Numerical Summaries {#numericalsummaries}

The numerical summary must match up with the type of variable(s).

|Variable|Type of summary|
|----------|------------------------|
|1 Qualitative|frequency table, most common category|
|1 Quantitative|mean, median, SD, IQR etc|
|2 Qualitative|contingency table|
|2 Quantitative|correlation, linear model|
|1 Quantitative, 1 Qualitative|mean, median, SD, IQR etc across categories|

<br>
We'll keep working with the `mtcars` dataset.
So again remind yourself what it is like.
```{r,eval = T}
str(mtcars)
dim(mtcars)
head(mtcars)
help(mtcars)
```

<br>

## Frequency and contingency tables

- A frequency table summarises **1** qualitative variable.

```{r,results='hide'}
table(mtcars$gear)
```

- A contingency table summarises **2** qualitative variable. 

```{r,results='hide'}
table(mtcars$gear, mtcars$am)
```

<br>

## Mean and median

- The mean and median measure centre for quantitative variables.
<br>
```{r,results='hide'}
mean(mtcars$gear)
median(mtcars$gear)
```

<br>

## Standard deviation (SD)

- The standard deviation measures spread for quantitative variables.

- The `sd` command calculates the **sample** standard deviation. The squared SD is the variance.

```{r,results='hide'}
sd(mtcars$gear)
sd(mtcars$gear)^2
var(mtcars$gear)
```

- The `popsd` command calculates the **population** standard deviation, but requires the `multicon` package.
<br>
```{r,results='hide', message=F, warning = F}
#install.packages(multicon)   # a package only needs to be installed once.
library(multicon)
popsd(mtcars$gear)

# Longer way
N = length(mtcars$gear)
sd(mtcars$gear)*sqrt((N-1)/N)
```

- Note: When we model a population by the box model [Section 8 and following], we will require the population SD.

<br>

## Interquartile range (IQR)

- The quickest method is to use `IQR`. 
```{r,results='hide'}
IQR(mtcars$gear)
```

- There are lots of different methods of working out the quartiles. We can use the `quantile` command, and then work out the IQR.
```{r,results='hide'}
quantile(mtcars$gear)
quantile(mtcars$gear)[4]-quantile(mtcars$gear)[2]
```
What is the 50% quantile equivalent to?

<br>

## Summary

- The numerical summaries for quantitative variables can all be produced with `summary`, which is an expanded version of the 5 number summary. Sometimes these values will vary from using `quantile` as there are different conventions for calculating quartiles.
```{r,results='hide'}
summary(mtcars$mpg)  # 1 variable
summary(mtcars)  # All variables
```

- We can consider a subset of the data.
Here, we choose the mpg of cars which have a weight greater or equal to 3.

```{r,results='hide'}
summary(mtcars$mpg[mtcars$wt>=3])
```

Here we take all the data from mtcars dataset for a specific cylinder e.g. 6. 

```{r,results='hide'}
mtcars[ which(mtcars$cyl==6), ]
```


<br>

## Linear Correlation

- We can consider the linear correlation between pairs of quantitative variables.
```{r,results='hide'}
cor(mtcars)
```
<br><br>