# Normal model {#normalmodel}

<br>

## Plot curve

- To plot a normal curve, we use `dnorm' for a range of $x$ values.

```{r,results='hide', fig.show='hide'}
par(mfrow=c(1,2)) # plots 2 plots on same row
curve(dnorm(x),-1,1)
curve(dnorm(x),-5,5)
```

When does the Normal curve start to disappear?

<br>

## Plot area under curve (Ext)

- To plot an area under the curve, we use `polygon`.

```{r,results='hide', fig.height=4, fig.show='hide'}
curve(dnorm(x), xlim = c(-2, 2),ylab="",axes=FALSE)
abline(h = 0)
sequence = seq(-2, 0.8, 0.1)
polygon(x = c(sequence,0.8,-2),
        y = c(dnorm(c(sequence)),0,0),
        col = "purple")
axis(1, at=c(-2,-1,0,0.8,1,2), pos=0)
```

<br>

## Add normal to a histogram

```{r,results='hide', fig.show='hide'}
hist(mtcars$wt,prob=T)
m=mean(mtcars$wt)
s=sd(mtcars$wt)
curve(dnorm(x, mean=m, sd=s), 
col="darkgreen", lwd=2, add=TRUE)
```

<br>

## Probabilities: Standard normal

### Lower tail

To calculate P(Z < 0.8) use `pnorm`.

```{r,results='hide'}
pnorm(0.8)
```

<br>

### Upper tail

To calculate P(Z > 0.8):

```{r,results='hide'}
pnorm(0.8,lower.tail=F)
1-pnorm(0.8)
```

<br>


### Interval

To calculate P(0.3 < Z < 0.7):

```{r,results='hide'}
pnorm(0.7)-pnorm(0.3)
```

<br>

## Probabilities: General normal

If $X \sim N(5, 4^2)$, then to calculate P(X < 8), we use `pnorm(8,mean,sd)`.

```{r,results='hide'}
pnorm(8,mean=5,sd=4)
pnorm(8,5,4)
```

<br>
