# Linear model {#linearmodel}

## Scatter plot

- To produce a scatter plot, use `plot(<x variable>, <y variable>)`.

```{r,results='hide', fig.show='hide'}
plot(mtcars$mpg,mtcars$wt)
plot(mtcars$mpg,mtcars$wt, xlab="mpg", ylab="weight", main="Scatter plot of weight vs mpg")
```

<br>

## Correlation coefficient

- To calculate the correlation coefficient, use `cor(<x variable>, <y variable>)`.


```{r,results='hide'}
cor(mtcars$mpg,mtcars$wt)
```

<br>

## Regression line

- To produce the regression line, use `lm(<y variable>~<x variable>)`.

```{r,results='hide'}
lm(mtcars$wt~mtcars$mpg)
#Or
L=lm(mtcars$wt~mtcars$mpg)
L$coeff
#Or
lm(wt~mpg,data=mtcars)
```

- Now you can add this line to the scatter plot.

```{r,results='hide', fig.show='hide'}
plot(mtcars$mpg,mtcars$wt)
abline(L)
```

<br>

## Residual plot 

- The residual plot helps to detect any pattern not captured by the linear model.

    - If it is a random scatter, then the linear model seems appropriate.
    - If it shows a pattern (eg quadratic), then another model should be considered.

- To produce a residual plot:

```{r,results='hide', fig.show='hide'}
plot(mtcars$mpg,L$residuals, xlab = "mpg", ylab = "Residuals")
abline(h = 0, col = "blue")
```