# Simulate chance {#simulatechance}

## Set seed

Setting a seed is a way to simulate the same results each time (eg for sharing with a collaborator).

- Notice what happens below if we don't set a seed.

```{r,results='hide'}
sample(c("H", "T"), 5, replace = T)
sample(c("H", "T"), 5, replace = T)
```

- Now set a seed.

```{r,results='hide'}
set.seed(1)  # ensures the same simulation, "1" can be whatever you choose
sample(c("H", "T"), 5, replace = T)
set.seed(1)
sample(c("H", "T"), 5, replace = T)
```

## Sample (Coin)

- To simulate 1 fair coin toss

```{r,results='hide'}
set.seed(1) 
sample(c("H", "T"), 1, replace = T)
```

- To simulate 1 biased coin toss

```{r,results='hide'}
set.seed(1)
sample(c("H", "T"), 1, replace = T,prob = c(0.9, 0.1))
```

- To simulate 10 fair coin tosses, and summarise chances in a table:

```{r,results='hide'}
set.seed(1)
sample1 = sample(c("H", "T"), 10,replace=T)
table(sample1)/10
```

- Alternatively, you could use
```{r,results='hide'}
set.seed(1)
replicate(10, sample(c("H", "T"),1,replace=T))
```

- Alternatively, you could use the Binomial *,
```{r,results='hide'}
set.seed(1)
rbinom(n = 10, 1, 0.5)
```

## Sample (Dice)

- To simulate 1 fair die toss

```{r,results='hide'}
set.seed(1)
sample(c(1:6), 1, replace = T)
```

- To simulate 1 biased die toss

```{r,results='hide'}
set.seed(1)
sample(c(1:6), 1, replace = T,prob = c(0.1,0.2,0.3,0.2,0.1,0.1))
```

- To simulate 10 fair dice tosses, and summarise chances in a table:

```{r,results='hide',fig.show='hide'}
set.seed(1)
sample(c(1:6), 10, replace = T)
table(sample(c(1:6), 10, replace = T))/10
barplot(table(sample(c(1:6), 10, replace = T))/10, xlab="dice face", ylab="chance")
```

- Alternatively, you could use
```{r,results='hide'}
set.seed(1)
table(replicate(10, sample(c(1:6), 10, replace = T)))
```

## Function **

To toss a die 10 times, and find the chance of getting a 6:

```{r,results='hide'}
tossdie = function() {
    rolls = sample(1:6, size = 10, replace = TRUE)
    condition = sum(rolls == 6) > 0
    return(condition)
}
# Simulate 10000 times
sim = replicate(10000, tossdie())
sum(sim)/length(sim)
```

<br> 

## Binomial Model *

Suppose we have 10 trials with P(event) = 0.3.

The chance of getting 4 events:
```{r,results='hide'}
dbinom(4,10,0.3)
```

The chance of getting less than or equal to 5 events:
```{r,results='hide'}
pbinom(5,10,0.3)
```

<br><br>